---
title: "Population Impact of Tornadoes and Economic Impact of Floods"
author: "Arun K Viswanathan"
date: "January 22, 2015"
output:
  html_document:
    theme: united
    toc: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Synopsis

This paper analyzes the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database. This database tracks characteristics of major storms and 
weather events in the United States, including when and where they occur, as 
well as estimates of any fatalities, injuries, and property damage. The paper
addresses two questions:

1. Across the United States, which storm event types have the most impact on 
human populations? The analysis shows that *tornado* events cause most 
fatalities and injuries across the United States.
2. Across, the United States, which storm event types have the greatest
economic consequences? The analysis shows that *flood* events cause the greatest
amount of economic damage across the United States.

All data analysis was performed with `R` and the code is included in this paper.
The `R` session information is shown below:

```{r session-info}
sessionInfo()
```

# Data Processing

The 
[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
was provided by Dr. Roger D. Peng, PhD as part of the
_[Reproducible Research](https://class.coursera.org/repdata-010)_ 
Coursera course. The data is described in 
_[National Weather Service Instruction 10-1605](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)_ 
from the NOAA. 

The code block below shows how the data is loaded and pre-processed before 
analysis. The data is first downloaded from the source and persisted in
the local directory `data`. The data is then uncompressed and loaded as a 
`dplyr` table data frame. Column names are corrected to be lower case and 
suitable for use in `R`. Only the columns relevant to this analysis are 
retained - `evtype`, `fatalities`, `injuries`, `propdmg`, `propdmgexp`, 
`cropdmg`, and `cropdmgexp`. The event type data in the `evtype` column is also 
corrected to match the closest type listed in section 2.1 
_Permitted Storm Data Events_ of the NOAA document. 

```{r load-data, message = FALSE, warning = FALSE}
library(dplyr)
library(lubridate)

# Download and read file
dataFileName <- "data/StormData.csv.bz2"
if (!file.exists(dataFileName)) {
    dir.create("data", showWarnings = FALSE)
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = dataFileName,
                  method = "curl")
    }
data <- tbl_df(read.csv(bzfile(dataFileName), na.strings = ""))

# Fix up column names
names(data) <- make.names(tolower(names(data)))

# Get year range in the data
data$bgn_date <- mdy_hms(data$bgn_date, truncated = 1)
yearRange <- range(year(data$bgn_date))

# Eliminate events with no fatalities, injuries or damage
# Keep only the columns of interest
data <- data %>%
    filter(fatalities != 0 | injuries != 0 | propdmg != 0 | cropdmg != 0) %>%
    select(evtype, fatalities, injuries, propdmg, propdmgexp, cropdmg, cropdmgexp)

# Fix typos in event names and equate to the valid set
data$evtype[which(data$evtype == "AVALANCE")] <- "AVALANCHE"
data$evtype[which(data$evtype == "STORM SURGE")] <- "STORM SURGE/TIDE"
data$evtype[which(data$evtype == "TORNDAO")] <- "TORNADO"
data$evtype[grep("cold", data$evtype, ignore.case = TRUE)] <- "COLD/WINDCHILL"
data$evtype[grep("dust", data$evtype, ignore.case = TRUE)] <- "DUST STORM"
data$evtype[grep("fld$", data$evtype, ignore.case = TRUE)] <- "FLOOD"
data$evtype[grep("flood", data$evtype, ignore.case = TRUE)] <- "FLOOD"
data$evtype[grep("fog", data$evtype, ignore.case = TRUE)] <- "DENSE FOG"
data$evtype[grep("freez", data$evtype, ignore.case = TRUE)] <- "FROST/FREEZE"
data$evtype[grep("hail", data$evtype, ignore.case = TRUE)] <- "HAIL"
data$evtype[grep("heat", data$evtype, ignore.case = TRUE)] <- "HEAT"
data$evtype[grep("hurricane", data$evtype, ignore.case = TRUE)] <- "HURRICANE/TYPHOON"
data$evtype[grep("ice", data$evtype, ignore.case = TRUE)] <- "ICE STORM"
data$evtype[grep("lightning", data$evtype, ignore.case = TRUE)] <- "LIGHTNING"
data$evtype[grep("rain", data$evtype, ignore.case = TRUE)] <- "HEAVY RAIN"
data$evtype[grep("rip ", data$evtype, ignore.case = TRUE)] <- "RIP CURRENT"
data$evtype[grep("snow", data$evtype, ignore.case = TRUE)] <- "HEAVY SNOW"
data$evtype[grep("surf", data$evtype, ignore.case = TRUE)] <- "HIGH SURF"
data$evtype[grep("thunder", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("tornado", data$evtype, ignore.case = TRUE)] <- "TORNADO"
data$evtype[grep("tropical", data$evtype, ignore.case = TRUE)] <- "TROPICAL STORM"
data$evtype[grep("tstm", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("typhoon", data$evtype, ignore.case = TRUE)] <- "HURRICANE/TYPHOON"
data$evtype[grep("wild", data$evtype, ignore.case = TRUE)] <- "WILDFIRE"
data$evtype[grep("wind", data$evtype, ignore.case = TRUE)] <- "HIGH WIND"
data$evtype[grep("winter", data$evtype, ignore.case = TRUE)] <- "WINTER STORM"

nobs <- format(nrow(data), scientific = FALSE, big.mark = ",")
evtypeCount <- length(unique(data$evtype))
```

After cleanup, the data contains `r nobs` observations of storm event data from 
`r yearRange[1]` to `r yearRange[2]`. Around `r evtypeCount` event types are 
available in the cleaed up data.

# Results

## Storm Event Types and Population 

The code below analyzes the total number of fatalities and injuries for
each event type.

```{r analyze-by-pop, message = FALSE, warning = FALSE}
# Summarize the fatalities and injuries for each storm event type
popImpactData <- data %>%
    select(evtype, fatalities, injuries) %>%
    group_by(evtype) %>%
    summarise(fatalities = sum(fatalities), injuries = sum(injuries)) %>%
    filter(fatalities > 0 | injuries > 0)

# Analyze the events with most fatalities
topFatalities <- arrange(popImpactData, 
                         desc(fatalities))[1:10, c("evtype", "fatalities")]
eventTypes <- as.character(topFatalities$evtype)
topFatalities$evtype <- factor(eventTypes, levels = eventTypes, ordered = TRUE)
topFatalEvent <- eventTypes[1]
topFatalEventCount <- format(topFatalities$fatalities[1], big.mark = ",")

# Analyze the events with most injuries
topInjuries <- arrange(popImpactData, 
                       desc(injuries))[1:10, c("evtype", "injuries")]
eventTypes <- as.character(topInjuries$evtype)
topInjuries$evtype = factor(eventTypes, levels = eventTypes, ordered = TRUE)
topInjuryEvent <- eventTypes[1]
topInjuryEventCount <- format(topInjuries$injuries[1], big.mark = ",")
```

The plot below shows the top 10 events with the most number of fatalities. It
can be clearly seen that `r topFatalEvent` events are the most dangerous to
human populations resulting in `r topFatalEventCount` fatalities over the years.

```{r plot-fatalities, message = FALSE, warning = FALSE}
library(ggplot2)

# Plot fatalities
ggplot(topFatalities) +
    geom_bar(stat = "identity", fill = "lightcoral",
             aes(x = evtype, y = fatalities)) +
    ggtitle("Top 10 Most Fatal Storm Events") +
    xlab("Storm Event Type") +
    ylab("Number of Fatalities") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10))
```

The plot below shows the top 10 events with the most number of injuries. It
can be clearly seen that `r topInjuryEvent` events are the most dangerous to
human populations resulting in `r topInjuryEventCount` injuries over the years.

```{r plot-injuries, message = FALSE, warning = FALSE}
# Plot injuries
ggplot(topInjuries) +
    geom_bar(stat = "identity", fill = "lightcoral",
             aes(x = evtype, y = injuries)) +
    ggtitle("Top 10 Most Injurious Storm Events") +
    xlab("Storm Event Type") +
    ylab("Number of Injuries") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10))
```

## Storm Event Types and Economic Consequences

The code below analyzes the total property and crop damage for each event type.
The code combines the `propdmg` and `cropdmg` fields, accounting for their
exponents specified in `propdmgexp` and `cropdmgexp`. For the purposes of this
analysis, only valid exponents (NA, `K`, `M`, and `B`) are used and all others
are ignored. 

```{r analyze-by-econ, message = FALSE, warning = FALSE}
econImpactData <- data %>%
    select(evtype, propdmg, propdmgexp, cropdmg, cropdmgexp) %>%
    filter((propdmg > 0 & propdmgexp %in% c("K", "M", "B", NA))
           & (cropdmg > 0 & cropdmgexp %in% c("K", "M", "B", NA))) 

econImpactData[which(is.na(econImpactData$propdmgexp)), "propdmgexpn"] <- 1
econImpactData[which(econImpactData$propdmgexp == "K"), "propdmgexpn"] <- 1000
econImpactData[which(econImpactData$propdmgexp == "M"), "propdmgexpn"] <- 1000000
econImpactData[which(econImpactData$propdmgexp == "B"), "propdmgexpn"] <- 1000000000

econImpactData[which(is.na(econImpactData$cropdmgexp)), "cropdmgexpn"] <- 1
econImpactData[which(econImpactData$cropdmgexp == "K"), "cropdmgexpn"] <- 1000
econImpactData[which(econImpactData$cropdmgexp == "M"), "cropdmgexpn"] <- 1000000
econImpactData[which(econImpactData$cropdmgexp == "B"), "cropdmgexpn"] <- 1000000000

econImpactData$dmg <- econImpactData$propdmg * econImpactData$propdmgexpn +
    econImpactData$cropdmg * econImpactData$cropdmgexpn 

econImpactData <- econImpactData %>%
    select(evtype, dmg) %>%
    group_by(evtype) %>%
    summarise(damages = sum(dmg))

topDamages <- arrange(econImpactData, desc(damages))[1:10, ]
eventTypes <- as.character(topDamages$evtype)
topDamages$evtype = factor(eventTypes, levels = eventTypes, ordered = TRUE)
topDamageEvent <- eventTypes[1]
topDamageAmt <- format(topDamages$damages[1], big.mark = ",")

```

The plot below shows the top 10 events with the most damage. It can be clearly 
seen that `r topDamageEvent` events have the most economic impact resulting in 
$`r topDamageAmt` of damages over the years.

```{r plot-damages, message = FALSE, warning = FALSE}
# Plot damages
ggplot(topDamages) +
    geom_bar(stat = "identity", fill = "lightcoral",
             aes(x = evtype, y = damages/1000000000)) +
    ggtitle("Top 10 Most Damaging Storm Events") +
    xlab("Storm Event Type") +
    ylab("Amount of Damage (billions of $)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10))
```

