---
title: "NOAA Storm Data Analysis"
author: "Arun K Viswanathan"
date: "January 20, 2015"
output: html_document
---

# Overview

This paper analyzes the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database. This database tracks characteristics of major storms and 
weather events in the United States, including when and where they occur, as 
well as estimates of any fatalities, injuries, and property damage.

# Storm Data

The [storm data][1] was provided by Dr. Roger D. Peng, PhD as part of the
_[Reproducible Research][2]_ Coursera course. The data is described in [this 
document][2] from the NOAA. 

```{r load-data,echo=FALSE,eval=TRUE,warning=FALSE}
dir.create("data", showWarnings = FALSE)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              "data/StormData.csv.bz2", 
              "curl")
```

# Appendix

## Code 

### Downloading and Loading Data

```{r echo=TRUE,eval=FALSE}
<<load-data>>
```


[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "Storm Data"
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "National Weather Service Instruction 10-1605"